Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-22 02-antecedents.qmd
ajustes formato

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [campbell_policy_2012; pierson_when_1993].

>>>>

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].

### 2025-05-22 04-analisis.qmd
ajustes formato

#knitreg(list(m13.1, m13.2,m14.1, m14.2))
```

@@

texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.79,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"),
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)", 
                                    "Var: individuals, merit effort cwc", 
                                    "Var: individuals, perception inequality cwc", 
                                    "Cov: individuals (Intercept), merit effort cwc",
                                    "Cov: individuals (Intercept), perception inequality cwc",
                                    "Cov: individuals, merit effort cwc, perception inequality cwc",
                                    "Var: Residuals",
                                    "Var: individuals, merit talent cwc", 
                                    "Cov: individuals (Intercept), merit talent cwc",
                                    "Cov: individuals, merit talent cwc, perception inequality cwc"),
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))

@@

p7 <- marginaleffects::plot_slopes(model = m9.2,
                             variables = "merit_talent_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Talent (BE)",
       fill = NULL,
       title = "(B) People are rewarded for their intelligence",
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)
       Confidence intervals at 95%") +
  theme_ggdist()

@@

texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)",

>>>>

#texreg(list(m13.1, m13.2,m14.1, m14.2))
```

@@

texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),caption.above = T,
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.80,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"),
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::texreg(list(m8.1,m8.2,m9.1,m9.2),
               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
               caption = NULL,
               caption.above = T,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)", 
                                    "Var: individuals, merit effort cwc", 
                                    "Var: individuals, perception inequality cwc", 
                                    "Cov: individuals (Intercept), merit effort cwc",
                                    "Cov: individuals (Intercept), perception inequality cwc",
                                    "Cov: individuals, merit effort cwc, perception inequality cwc",
                                    "Var: Residuals",
                                    "Var: individuals, merit talent cwc", 
                                    "Cov: individuals (Intercept), merit talent cwc",
                                    "Cov: individuals, merit talent cwc, perception inequality cwc"),
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))

@@

p7 <- marginaleffects::plot_slopes(model = m9.2,
                             variables = "merit_talent_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Talent (BE)",
       fill = NULL,
       title = "(B) People are rewarded for their intelligence",
       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4. Confidence intervals at 95%") +
  theme_ggdist()

@@

texreg::texreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               caption.above = T,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.7,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)",

### 2025-05-22 paper.qmd
ajustes formato

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvardquarto.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
